---
title: "36-315 Final Project"
author: "Lawrence Jang, Jaemin Lee, Deepro Hoque"
date: "12/4/2021"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

For our 36-315 final project, our group decided to utilize Sean Lahman's
Baseball Database. Lahman's database can be installed using the command
`install.packages("Lahman")` and can be accessed by running
`library("Lahman")` in the R console. 

As our group is full of avid baseball fans, we wanted to use the Lahman 
database to answer a handful of research questions; our 3
research questions are listed below.

**1) How have the outcomes of a baseball game changed over time, with special**
**regard to the three true outcomes?**

**2) What background factors are significant or highly correlated with**
**producing successful Major League Baseball players?**

**3) What factors determine a winning MLB team, and how much does it impact**
**revenue and front office budgetary decisions when constructing an MLB team?**


## Lahman Database 

Before we dive into our research questions, we will provide some context
on the Lahman Database and the data we are working with. The Lahman Database
is compiled by an investigative reporter named Sean Lahman, who works for
USA Today. He refers to the database as "an open-source collection of baseball 
statistics".

The Lahman Database contains 25+ dataframes, which contain the cumulative
pitching, hitting, fielding, teams, and award statistics for MLB
from the different eras of baseball spanning 1871-1875. If you
would like more information on the database and its different tables, 
feel free to run the command `help(Lahman)`.

```{r, echo = FALSE, message=FALSE, warning=FALSE}
library(Lahman)
```

Thus, given the plethora of datasets and different fields to explore
in the Lahman database, we chose specific tables to tackle each question. We 
will go into detail on the dataframes we used for oour specific research
questions in their respective sections below.

## First Research Question

As stated above, our first research question was **How have the outcomes of a **
**baseball game changed over time, with special regard to the three true outcomes?**
The three true outcome trend signifies the change in baseball as the highest
probabilities ofat-bat outcomes have become walks, home-runs, and 
strikeouts over time. We attempted to investigate this change through 
time-series analysis.

For this research question, investigating the trends of at-bat outcomes, 
we pulled our data from the Batting table. The Batting table consists of batting 
statistics, with each row in the table representing a single player given a
certain year's different hitting metrics. The important variables
we utilized are the following:

+ playerID, an individual player ID code
+ yearID, Year
+ lgID, League ID
+ AB, number of at bats a batter accumulated in a year
+ H, number of hits by a batter in a year
+ X2B, number of doubles by a batter in a year
+ X3B, number of triples by a batter in a year
+ HR, number of home runs by a batter in a year
+ SB, number of stolen bases by a batter in a year
+ BB, number of a walks by a batter in a year
+ SO, number of strikeouts by a batter in a year

We used the variables above to aggregate league-wide trends through a 
time-series analysis to investigate the changes in the outcomes in baseball.

You can access the Batting table after loading in the Lahman Database as such 
below. We display the first five rows of Batting below.

```{r, echo=FALSE}
Batting[1:5,]
```

We performed filtering such that our data consists of statistics from the year 2000 and on, 
as this three true outcome trend has been a recent development over the 
past couple of decades. We decided to perform preprocessing to get
league-wide aggregates for our desired statistics. Furthermore, we decided to 
group by different leagues across the different years. 

```{r,warning=FALSE,echo=FALSE, message=FALSE}
library(tidyverse)
df = Batting
df = df %>% filter(yearID >= 2000)
batcounts = df %>% group_by(yearID, lgID) %>% dplyr::summarize(hr = sum(HR),sb = sum(SB), ab = sum(AB),xb = sum(X2B) + sum(X3B), h = sum(H), so = sum(SO), bb = sum(BB))
batcounts$hrate = batcounts$hr/batcounts$ab
batcounts$sb = batcounts$sb/batcounts$ab
batcounts$xbrate = batcounts$xb/batcounts$ab
batcounts$avg = batcounts$h/batcounts$ab
batcounts$krate = batcounts$so/batcounts$ab
batcounts$walkrate = batcounts$bb/batcounts$ab
batcounts$avg = batcounts$h/batcounts$ab
```

We display the first five rows of our newly constructed dataframe with the 
outcome rates below.

```{r, echo=FALSE}
batcounts[1:5,]
```

From our newly constructed statistics, we made the following plots below.

### Plot 1

Our first plot displays the historical walk and strikeout rates from the 
last 20 MLB seasons. We have a geom_line using the `loess` method on 
our ggplot for both walk and strikeout rates, and we see a stark increase
in the number of strikeouts over the past 20 years and a slight increase
in walks over the past 5 years. This graph shows that the trend of the
true three outcomes is increasing and the everchanging nature of baseball
outcomes with different eras.

```{r, echo=FALSE}
ggplot(batcounts, aes(yearID)) + 
  geom_line(aes(y = krate, colour = "Strikeout Rate")) + 
  geom_line(aes(y = walkrate, colour = "Walk Rate")) + 
  labs(title = "MLB Historical Walk and Strikeout Rate", subtitle = "From the 2000-2020 Seasons") + 
  labs(y = "Percentage", x = "Year", colour = "Statistic") +
  facet_grid(~ lgID)
```

### Plot 2

Our second plot displays the historical home run and extra base hit rates from the 
last 20 MLB seasons. We again have a geom_line using the `loess` method on 
our ggplot for both home run and extra base hit rates. We see that there is a 
deep increase in the home run rate and a decrease in extra base hit rate in 
progressing years. This graph further adds to our three true outcome hypothesis;
we see that there is an intuitively inverse relationship between
extra base hits and home runs. As batters are emphasizing on hitting 
hard hit balls over the wall rather than throughout the field, this
has led to less balls being in play, with more home runs.

```{r, echo=FALSE}
ggplot(batcounts, aes(yearID)) + 
  geom_line(aes(y = hrate, colour = "Home Run Rate")) +
  geom_line(aes(y = xbrate, colour = "Extra Base Hits Rate")) + 
  labs(title = "MLB Historical Home Run and Extra Base Hit Rate", subtitle = "Beginning from the 2000 Season") + labs(y = "Percentage", x = "Year", colour = "Statistic") +
    facet_grid(~ lgID)
```

### Plot 3

For our final plot regarding our first research question, we decided to plot a 
time-series graph of the historical trend in MLB-wide average, split and
colored across the different leagues. This scatterplot with `loess` trend
lines shows that the average rate, which is the rate at which batters 
attain a hit from an at-bat, has progressivle decreased over the past 20 years.
We also see that there is a difference across the two leagues; as the AL 
(American League) has a designated hitter bat for the pitcher, this leads
to the inflation in average for the AL trend line. This graph shows again that
the ball is being put in play less in the MLB, and moving towards the 
three true outcomes.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = batcounts, aes(x = yearID, y = avg, col = lgID)) + geom_point() +
  geom_smooth(se = F) +
  labs(title = "MLB Historical Average Rate", subtitle = "Colored by Different Leagues",
       x = "Year", y = "Percentage")
```



## Second Research Question

Our second research question was **What background factors are significant or highly correlated with producing successful Major League Baseball players?**
As we are given biographical data consisting of player's birthdays, hometowns,
and home countries, we wanted to see if there were any consistent trends 
in producing MLB players.

For this research question, we pulled from the the People table. The People
table consists of biographical data for every player to play in the MLB.
We used the following variables from the People table to conduct our analysis.

+ playerID, an individual player ID code
+ birthMonth, Month player was born
+ birthCountry, Country where player was born
+ birthState, State where player was born

Here is a look into the first five rows of the People dataframe.

```{r, echo=FALSE}
People[1:5,]
```

For this research question, we employed the use of heat maps for the United 
States. We use `map_data("state")` to get the longitudinal and latitudinal 
values to create our heat map across different states. We display this 
the first 5 rows of this dataframe below. We also load in the 
`usdata` library in order to conver the state abbreviations to match
the region state names in the dataframe below.
```{r, echo=FALSE}
map_data("state")[1:5,]
```

### Plot 1

For our first plot, we made a bar plot displaying the frequencies of
the top 8 countries that produce the most MLB players. From this graph,
we get an understanding of the global scene on baseball players; most
MLB players are from the USA, followed by countries in Latin America.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
countries = People %>% group_by(birthCountry) %>% dplyr::summarize(n = n()) %>% arrange(-n)
countries = countries %>% top_n(8)
ggplot(data = countries, aes(x = reorder(birthCountry,-n), y = n)) + geom_bar(stat = "identity", fill = "blue", col = "orange") +
  labs(title = "MLB Players by Top 8 Home Countries", x = "Home Country", y = "Player Count")
```

### Plot 2

As we saw that most players are from the United States, we want to get
an understanding of which regions of the United States produce the most players.
Thus, we create a heat map of the United States, which the color of each state
representing the frequency of players produced by the respective state. We see
that the majority of players come from the state of California, with
a little more production of players in the Northeast, Midwest, and Southeast.

```{r, warning=FALSE, message=FALSE, echo=FALSE}
library(usdata)
state.data = map_data("state")
# we exclude Hawaii and Alaska for our map
us = subset(People, birthState != "AK" & birthState != "HI" & birthCountry == "USA")
us = us %>% group_by(birthState) %>% dplyr::summarize(n = n()) %>% arrange(-n)
us$lowerState = tolower(abbr2state(us$birthState))
# we merge the state locational data and player production datasets
merged = state.data %>% left_join(us, by = c("region" = "lowerState"))
ggplot(merged) +
  geom_polygon(aes(x = long, y = lat, group = group, fill = n), 
               color = "black") + 
  scale_fill_gradient2(high = "red", low = "darkblue", 
                       mid = "lightblue", midpoint = 100) +
  theme_void() +
  coord_map("polyconic") + 
  labs(title = "MLB Players by Home State",
       fill = "Player Count") +
  theme(plot.title = element_text(hjust = 0.5))
```

### Plot 3

For our third plot, we created a heat map based on the number of Hall of
Famers by state. By creating this graph, we will inspect if the states that
produce the number of truly successful players is different than the frequency
at which they produce all MLB players.
In order to do so, we accessed the the HallOfFame table
in the Lahman Database. By filtering the HallOfFame table
for `inducted ==  Y`, we get a dataframe of all the Hall of Fame members.

```{r, echo=FALSE}
inducted = HallOfFame %>% filter(inducted == "Y")
inducted[1:5,]
```

In order to create our heat map, we merge the inducted Hall of Fame
table with the player biographical table by playerID. Then, we follow the same
procedure as in Plot 2 to find the number of players by state by joining
with our longitudinal and latitudinal table.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# join the two tables
merge2 = merge(inducted, People, by = "playerID")
merge2 = merge2 %>% group_by(birthState) %>% dplyr::summarize(n = n()) %>% arrange(-n)
merge2$lowerState = tolower(abbr2state(merge2$birthState))
merge3 = state.data %>% left_join(merge2, by = c("region" = "lowerState"))
ggplot(merge3) +
  geom_polygon(aes(x = long, y = lat, group = group, fill = n), 
               color = "black") + 
  scale_fill_gradient2(high = "red", low = "blue", 
                       mid = "lightyellow1", midpoint = 15) +
  theme_void() +
  coord_map("polyconic") + 
  labs(title = "Hall Of Fame Players by Home State", subtitle = "States with 0 HOFers are Colored Grey",
       fill = "Player Count") +
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5))

```

In the plot, we see that the heat map loosely follows the heat map from 
Plot 2; thus, we can observe that the same top states California, New York,
and others in the Midwest and Northeast produce the MLB's top talent.


## Third Research Question

Our third research question we posed was **What factors determine a winning MLB team, and how much does it impact revenue and front office budgetary decisions when constructing an MLB team?**

In order to investigate this question, we utilized the Salaries and Teams 
tables given to us in the Lahman Database.

From the Salaries table, we find the player salary data for a given 
year and team. We display the first five rows of the table below.

```{r, echo=FALSE}
Salaries[1:5,]
```

From the Teams table, we are given the performance statistics and standings
of a team given a certain year. From this table, there are 48 variables;
however, we will mainly utilize the following variables:

+ yearID, the code of a specific year
+ lgID, the code of a given league
+ G, number of games played
+ W, number of wins
+ R, number of runs scored
+ RA, number of runs allowed

We want to utilize the mentioned variables to calculate the winning percentage
and run differential of a team for a given year. Given the study of the 
Pythagorean Theorem of Baseball (https://www.baseball-reference.com/bullpen/Pythagorean_Theorem_of_Baseball#:~:text=The%20Pythagorean%20Theorem%20of%20Baseball,a%20team's%20actual%20winning%20percentage.),
we want to use run difference as an estimate of predicting winning percentage.

```{r, echo=FALSE}
teams <- Teams %>% 
  filter(yearID >= 1901 & lgID %in% c("AL", "NL")) %>%
  group_by(yearID, teamID) %>%
  mutate(WinPct = W/G,
         rd = R - RA,
         proj = ((R^2)/(R^2 + RA^2)))
```

With these statistics, we create our first plot.

## Plot 1

We create a scatterplot with a trend line, with x = Run Differential and 
y = Winning Percentage for all the teams in our dataset from the 
year 1901. We see a clear positive correlation and relationship between
run differential and winning percentage, which suggest that the higher your run differential, it is more likely your winning percentage is higher. We can see
that run differential is a strong linear indicator of winning percentage,
regardless of league.

```{r, echo=FALSE, warning=FALSE,message=FALSE}
ggplot(teams, aes(x = rd, y = WinPct, col = lgID)) +
   geom_point(size = 0.5, col = "black") +
   geom_smooth(method="loess") + 
   labs(title = "Historical Run Differential vs. Winning Percentage", 
        subtitle = "Colored by League",
        x = "Run Differential", y = "Win percentage")
```

## Plot 2

Now, we want to analyze the relationship between team salary and a team's 
winning percentage, from the point of an MLB front office. We ask ourselves
the question: "How much money does it take to build a competent team?".

In order to get team salary data, we conduct the following processing. We
get the total salary for each team in our database, and then join it with
our Teams table to get the performance statistics and standings for
each team. We filter for teams beginning from the year 1985 to 
restrict to the modern era of ownership.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
#get team salaries
teamsal <- Salaries %>%
                group_by(yearID, teamID) %>%
                dplyr::summarize(Salary = sum(salary, na.rm = TRUE)) %>%
                select(yearID, teamID, Salary)
#join team salaries with budget
teamsSal <- teams %>%
                filter(yearID >= 1985) %>%
                left_join(teamsal, by = c("yearID", "teamID")) %>%
                select(yearID, teamID, attendance, Salary, WinPct, lgID) %>%
                as.data.frame(.)
```

Furthermore, we want to adjust for inflation in each decade. Thus, we will
make a decade variable that we will facet on in our plot. We also filter
to exclude the year 2020, as the database lacks the full data for the season
that was changed by the COVID-19 pandemic.

```{r, echo=FALSE}
teamsSal = teamsSal %>% filter(yearID < 2020)
teamsSal$decade = (teamsSal$yearID - (teamsSal$yearID %% 10))
```

We plot our Winning Percentage vs Salary plot below, facetted on each decade.
We see that with each increasing decade, there is greater parity in salary
and a more positive relationship with winning percentage and a team's salary. 
However, from a front office point of view, we see that although more 
money typically leads to more success, there are many teams in the 2000s
and 2010s decades that had a high winning percentage with a low salary.
Thus, keeping a balance is ideal and possible in the world of managing 
baseball teams!

```{r, echo=FALSE, warning=FALSE, message=FALSE}
ggplot(teamsSal, aes(x = WinPct, y = Salary, col = lgID)) +
   geom_point(size = 0.5, col = "Black") +
   geom_smooth(method="loess", se = F) + 
   labs(title = "Historical Salary vs. Winning Percentage", 
        subtitle = "Facetted on Each Decade",
        x = "Win Percentage", y = "Salary") +
  facet_grid(~ decade)
```

## Conclusion

From our three research questions, we were able to answer and tell a story
with the use of data and statistical visualizations. From the Lahman Database,
we were able to investigate a time series analysis of the changing outcomes of 
a baseball game, the geographical hot spots of producing MLB players,
and the importance of run differential and team salary on a team's success.
We truly enjoyed working with this database and we hope that you
enjoyed our analysis.
Thank you and feel free to reach out if you have any questions!